---
title: "Getting started with rmnchfunding"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with rmnchfunding}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rmnchfunding)
```

## Why this package exists

No OECD dataset reports how much aid goes to reproductive, maternal, newborn
and child health. The CRS records disbursements against purpose codes that
are broader than RMNCH — "basic health care", "basic nutrition" — and a
provider's contribution to the WHO or the Global Fund appears as a single
core payment with no sectoral breakdown at all.

The Muskoka method closes both gaps with coefficients: a fixed share of each
purpose code is attributed to the universe of interest, and a fixed share of
each multilateral agency's spending is attributed alongside it. This package
holds those coefficients, fetches the two OECD sources they apply to, and
does the arithmetic.

## Status

Under construction. The coefficient tables are in place; the OECD fetchers
and `muskoka()` itself are not yet written, so there is nothing here that
produces an estimate.

## The three universes

Every coefficient is given for three nested-in-spirit but separately
estimated universes:

| Universe | Scope |
|---|---|
| `rmnch` | Reproductive, maternal, newborn and child health |
| `srhr` | Sexual and reproductive health and rights |
| `fp` | Family planning |

`muskoka()` will take one of these as its `universe` argument.

## The coefficient tables

Bilateral flows are weighted by CRS purpose code:

```{r}
head(sector_weights[sector_weights$universe == "rmnch", ], 8)
```

Multilateral spending is weighted by agency, and needs two keys rather than
one:

```{r}
agency_weights[
  agency_weights$agency == "Global Fund" & agency_weights$universe == "rmnch",
]
```

`data_year` is the year of the *spending*: each weight is the proportion of
that agency's own disbursements in that year benefiting the universe, so it
moves as the agency's disbursement mix moves.

`report_edition` is the year of the *report*. Every edition recomputes the
weights for all the years it covers, including years an earlier edition
already published, as the underlying data is revised. Those revisions are
large enough to matter:

```{r}
adb <- agency_weights[
  agency_weights$agency == "Asian Development Bank" &
    agency_weights$universe == "rmnch" &
    agency_weights$data_year %in% 2022:2023,
]
adb[order(adb$data_year, adb$report_edition), ]
```

The 2023 weight is 5.18% as first published and 13.42% a year later. So
**take the weights and the published totals from the same edition** — mixing
them reproduces neither report. The editions also differ in price base: 2022
constant prices in 2025, 2023 constant prices in 2026.

Between them the two editions cover four spending years, 2021 to 2024, which
is what makes the four per-donor RMNCH weights recoverable — four unknowns
need four independent published years, and neither edition alone has them.

## Weights that are missing on purpose

Some coefficients are `NA`, and that is not the same as zero:

```{r}
na_cells <- sector_weights[is.na(sector_weights$weight), ]
table(na_cells$universe)
```

All four are RMNCH weights — malaria control, tuberculosis control, STD
control including HIV/AIDS, and general budget support. The source marks
these `varies*` because their RMNCH share is set **per donor country**, and a
table with one weight per code cannot hold a per-donor value. They will move
to their own donor-keyed table once derived. The SRHR and family-planning
universes are complete. See `?sector_weights`.

## Where the coefficients come from

Both tables are transcribed from the Donors Delivering for SRHR Report 2026,
pages 110–111. RMNCH follows the Muskoka 2 methodology developed by the London
School of Hygiene and Tropical Medicine, family planning follows the revised
Muskoka methodology agreed at the 2012 London Summit, and SRHR follows the
Donors Delivering methodology.

The three universes overlap by construction. Adding a donor's RMNCH, SRHR and
FP figures together double-counts, and the report says so explicitly — they
are meant to be read separately.

Because a missing weight is unknown rather than zero, `muskoka()` will refuse
to compute a total that depends on one instead of quietly treating it as
nothing — a zero here would understate the result without leaving a trace.

## What this vignette is not

Reference documentation. That is generated from the roxygen comments and
reached with `?sector_weights` and `?agency_weights`. A vignette explains
*when and why*; the help page explains *what and how*.
