---
title: "Getting started with rmnchfunding"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with rmnchfunding}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rmnchfunding)
```

## Why this package exists

No OECD dataset reports how much aid goes to reproductive, maternal, newborn
and child health. The CRS records disbursements against purpose codes that
are broader than RMNCH — "basic health care", "basic nutrition" — and a
provider's contribution to the WHO or the Global Fund appears as a single
core payment with no sectoral breakdown at all.

The Muskoka method closes both gaps with coefficients: a fixed share of each
purpose code is attributed to the universe of interest, and a fixed share of
each multilateral agency's spending is attributed alongside it. This package
holds those coefficients, fetches the two OECD sources they apply to, and
does the arithmetic.

## Status

Under construction. The coefficient tables and both OECD fetchers are in
place. `muskoka()` itself is not written yet, and four per-donor RMNCH weights
are still to be derived, so the package does not produce a Muskoka estimate —
but `oecd_crs()`, `oecd_multi()` and `crs_classify()` are usable on their own.

The chunks that reach the OECD API are shown but not executed here, so that
building this vignette never depends on a network connection. Figures quoted
in the text are from the United States in 2022, in constant 2023 prices.

## The three universes

Every coefficient is given for three nested-in-spirit but separately
estimated universes:

| Universe | Scope |
|---|---|
| `rmnch` | Reproductive, maternal, newborn and child health |
| `srhr` | Sexual and reproductive health and rights |
| `fp` | Family planning |

`muskoka()` will take one of these as its `universe` argument.

## The coefficient tables

Bilateral flows are weighted by CRS purpose code:

```{r}
head(sector_weights[sector_weights$universe == "rmnch", ], 8)
```

Multilateral spending is weighted by agency, and needs two keys rather than
one:

```{r}
agency_weights[
  agency_weights$agency == "Global Fund" & agency_weights$universe == "rmnch",
]
```

`data_year` is the year of the *spending*: each weight is the proportion of
that agency's own disbursements in that year benefiting the universe, so it
moves as the agency's disbursement mix moves.

`report_edition` is the year of the *report*. Every edition recomputes the
weights for all the years it covers, including years an earlier edition
already published, as the underlying data is revised. Those revisions are
large enough to matter:

```{r}
adb <- agency_weights[
  agency_weights$agency == "Asian Development Bank" &
    agency_weights$universe == "rmnch" &
    agency_weights$data_year %in% 2022:2023,
]
adb[order(adb$data_year, adb$report_edition), ]
```

The 2023 weight is 5.18% as first published and 13.42% a year later. So
**take the weights and the published totals from the same edition** — mixing
them reproduces neither report. The editions also differ in price base: 2022
constant prices in 2025, 2023 constant prices in 2026.

Between them the two editions cover four spending years, 2021 to 2024, which
is what makes the four per-donor RMNCH weights recoverable — four unknowns
need four independent published years, and neither edition alone has them.

## Weights that are missing on purpose

Some coefficients are `NA`, and that is not the same as zero:

```{r}
na_cells <- sector_weights[is.na(sector_weights$weight), ]
table(na_cells$universe)
```

All four are RMNCH weights — malaria control, tuberculosis control, STD
control including HIV/AIDS, and general budget support. The source marks
these `varies*` because their RMNCH share is set **per donor country**, and a
table with one weight per code cannot hold a per-donor value. They will move
to their own donor-keyed table once derived. The SRHR and family-planning
universes are complete. See `?sector_weights`.

## Where the coefficients come from

Both tables are transcribed from the Donors Delivering for SRHR Report 2026,
pages 110–111. RMNCH follows the Muskoka 2 methodology developed by the London
School of Hygiene and Tropical Medicine, family planning follows the revised
Muskoka methodology agreed at the 2012 London Summit, and SRHR follows the
Donors Delivering methodology.

The three universes overlap by construction. Adding a donor's RMNCH, SRHR and
FP figures together double-counts, and the report says so explicitly — they
are meant to be read separately.

Because a missing weight is unknown rather than zero, `muskoka()` will refuse
to compute a total that depends on one instead of quietly treating it as
nothing — a zero here would understate the result without leaving a trace.

## Fetching the two OECD sources

`oecd_crs()` pulls a donor's bilateral disbursements for the weighted purpose
codes, broken down by recipient and year:

```{r, eval = FALSE}
crs <- oecd_crs("USA", years = 2022, prices = "constant", base = 2023)
```

`oecd_multi()` pulls the other half — core contributions to the eleven weighted
multilateral agencies:

```{r, eval = FALSE}
mul <- oecd_multi("USA", years = 2022, prices = "constant", base = 2023)
```

Two arguments deserve attention.

**`prices` is never implicit.** `"current"` leaves each year in its own prices.
`"constant"` requires a `base` year, because OECD rebases its own constant
series with each release — 2024 at the time of writing, 2023 a few months
earlier — so a default would change meaning underneath you between releases.
Values are always fetched in current prices and deflated here with OECD's
per-donor deflators, which is what allows any base year to be requested. To
reproduce a published Donors Delivering figure, match its edition: base 2022
for the 2025 edition, base 2023 for the 2026 edition.

**`recipients` controls de-duplication.** A CRS query with the recipient
dimension open returns aggregates and countries as sibling rows, so
"Developing countries", "Africa" and "Kenya" all come back together. Summing
them counts Kenya four times. The default `"countries"` keeps only
non-aggregate rows and is the only setting whose rows can safely be summed.
`"all"` keeps everything, which is what `crs_classify()` needs.

Note that de-duplicating is not the same as reducing to countries. OECD
reports the part of a donor's spending it cannot attribute to any country in
`_X` buckets — "Developing countries unspecified", "Sub-Saharan Africa
unspecified". These have no members, so they are correctly kept and summed
alongside countries, but they are not countries: for the United States in 2022
they are 42% of the total. They carry `is_unallocated` so you can sum either
way.

## Classifying recipients

OECD classifies recipients several ways at the same time, and this is easy to
get wrong: the recipient codelist has nineteen top-level groupings that look
comparable in a flat list but are not. `crs_classify()` returns one
*classification* at a time, each of which sums to the same grand total.

```{r, eval = FALSE}
d <- oecd_crs("USA", years = 2022, prices = "constant", base = 2023,
              recipients = "all")

crs_classify(d, "geographic")
crs_classify(d, "dac_income")
crs_classify(d, "wb_income")
```

For the United States in 2022, all three come to the same 24,281.9 million:

| Geographic | | DAC income | | World Bank income | |
|---|---:|---|---:|---|---:|
| Africa | 9,817.6 | *unallocated* | 10,089.9 | *unallocated* | 10,089.9 |
| *unspecified* | 7,766.8 | Least developed | 8,759.5 | Low income | 6,506.8 |
| Asia | 4,164.2 | Lower-middle | 3,018.9 | Lower-middle | 4,352.9 |
| America | 1,435.2 | Upper-middle | 1,861.4 | Upper-middle | 2,034.4 |
| Europe | 890.0 | Other low income | 552.1 | Not classified | 1,274.4 |
| Oceania | 208.0 | | | High income | 23.5 |
| **24,281.9** | | **24,281.9** | | **24,281.9** | |

Three things follow that are worth knowing before using these figures.

**"Developing countries" is the total, and "least developed countries" is not
a separate universe.** `DPGC` is the grand total. `LDC` is a *tier* of the DAC
income classification — the same money cut a different way — so adding it to a
geographic figure double counts.

**The two income classifications are genuinely different.** The DAC List of ODA
Recipients has a least-developed tier and no high-income group; the World Bank
classification has the reverse, plus a group for countries it does not
classify at all.

**Some groupings are not classifications.** `HIPC`, `LLDC`, `SIDS`, `FSCAC` and
`ACP` are overlapping flags a country carries *in addition* to its place in
every classification above — Ethiopia is in Africa, Sub-Saharan Africa,
Eastern Africa, least developed, landlocked, World Bank low income, heavily
indebted and fragile simultaneously. They do not partition anything (`FSCAC`
alone is 47% of the United States total), so `scheme` cannot name them.

## Levels

Each classification can be cut at a chosen depth, and every level still sums
to the same total:

```{r, eval = FALSE}
crs_classify(d, "geographic", level = "continent")   # the default, level 1
crs_classify(d, "geographic", level = "subregion")
crs_classify(d, "dac_income", level = "country")
```

| Classification | Levels |
|---|---|
| `geographic` | `total`, `continent`, `region`, `subregion`, `country` |
| `dac_income` | `total`, `tier`, `country` |
| `wb_income` | `total`, `group`, `country` |

A level is a **frontier**, not a slice at a fixed depth. Branches that have
already ended stand in for themselves, because the hierarchy is ragged: Africa
divides into regions and then subregions, while Europe holds its countries
directly. So the geographic `"region"` level contains African regions and
European countries side by side. Taking "every node at depth two" instead would
drop every European country, and the level would no longer add up.

The same applies to unallocated spending, which enters at whatever level it was
reported. A country-level cut therefore still contains regional residuals such
as "Sub-Saharan Africa unspecified"; those rows are marked `is_unallocated`.

`crs_classify()` verifies on every call that the parts sum to OECD's own
reported total, and fails rather than return parts that do not make a whole.
That is not defensiveness for its own sake: OECD's published hierarchy does not
always describe its published aggregates. `INC_X` includes `DPGC_X` in the data
but does not list it as a child in the codelist, which cost exactly 7,766.8
million — 32% of this donor's total — the first time an income classification
was cut to country level.

## Which countries are imputed, and from where

Four RMNCH weights vary by recipient and year rather than being global
constants (12262 malaria, 12263 tuberculosis, 13040 STD including HIV/AIDS,
51010 general budget support). They are computed from IHME Global Burden of
Disease case counts and World Bank health-expenditure and population series.

Not every OECD recipient appears in those sources. Small territories such as
Wallis and Futuna, and OECD programmes such as the Mekong Delta, have no
disease burden or expenditure data of their own. Rather than leave them
without a weight, their geographic group's figure is substituted, trying the
narrowest group first: subregion, then region, then continent.

That figure is a **burden-weighted** mean of the group's members — weighted by
all-age case counts for the disease codes, and by population for general
budget support. This is not a stylistic choice: a ratio of summed cases across
a group is identically a mean of member ratios weighted by their denominators,
so weighting this way makes a substituted value equal what an aggregate of the
source data over that group would report. An unweighted mean would let a
country with almost no tuberculosis count as heavily as one carrying most of
the region's cases; for the Caribbean the two differ by 38%.

Every weight records how it was obtained, so a substituted value is never
mistaken for an observed one:

```{r}
table(rmnch_recipient_weights$source)
```

The substitution applies to each component separately, not just the total, so
`rh + mnh + ch` still equals `weight` exactly. For a recipient with no data of
its own:

```{r}
wlf <- rmnch_recipient_weights[
  rmnch_recipient_weights$recipient_code == "WLF" &
    rmnch_recipient_weights$year == 2023,
]
wlf[c("purpose_code", "rh", "mnh", "ch", "weight", "source")]
```

Note the malaria row: `mnh` keeps its fixed 0.15 while `ch` is 0, because
Polynesia is malaria-free. Averaging all three components rather than assuming
`mnh` is zero is what keeps the identity intact.

## The crosswalk, as a file

Three sources name countries three different ways, and none is reachable from
another by plain string matching:

| | Cote d'Ivoire |
|---|---|
| OECD | `Côte d’Ivoire` (circumflex, curly apostrophe) |
| World Bank | `Cote d'Ivoire` (no circumflex, ASCII apostrophe) |
| GBD | `Côte d'Ivoire` (circumflex, ASCII apostrophe) |

`recipient_map()` returns all three spellings together with each recipient's
geographic ancestors and, for each of the four varying purpose codes, whether
its weight is its own or borrowed:

```{r}
m <- recipient_map()
m[m$recipient_code == "CIV",
  c("recipient_name", "wb_name", "gbd_location_name")]
```

Ask it directly which recipients borrow a weight, and from which level:

```{r}
recipient_map("12262", imputed_only = TRUE)[
  c("recipient_name", "continent_name", "source")
]
```

The same content ships as a plain file, for checking the method without
running R:

```{r, eval = FALSE}
read.csv(system.file("extdata", "recipient_crosswalk.csv",
                     package = "rmnchfunding"))
```

Note `region_name` is `NA` for European recipients: the DAC hierarchy is
ragged, and Europe is the one continent it does not subdivide. The geography
throughout is OECD DAC, not UN M49 and not the World Bank's regions — they
disagree about Turkiye, Egypt and Central Asia, among others.

## Why the malaria constant is unconditional

Malaria's weight is `0.15 + CH`, where `CH` is the under-5 share of malaria
cases and `0.15` is a fixed allowance for malaria in pregnancy taken from the
Countdown method. The constant applies to **every** country and year: all 1,232
malaria MNH cells in the published reference are exactly 0.15, with no regional
or endemicity condition.

That has a consequence worth stating plainly. GBD 2023 reports **no malaria at
all** in 118 of its 204 locations, so for those recipients `CH = 0` and the
entire weight is the constant. A malaria-control disbursement to a malaria-free
country is still credited 15% to maternal and newborn health.

It is reasonable to ask whether the constant should instead be zero where there
is no burden. That variant is easy to construct from the shipped weights, since
a zero `ch` identifies exactly those recipient-years:

```{r}
w <- rmnch_recipient_weights[rmnch_recipient_weights$purpose_code == "12262", ]

# variant B: drop the MNH constant where no malaria burden is recorded
w$weight_variant_b <- ifelse(w$ch == 0, 0, w$weight)

table(
  `MNH applied` = ifelse(w$ch == 0, "no burden recorded", "burden recorded"),
  `variant B weight` = ifelse(w$weight_variant_b == 0, "0", "0.15 + CH")
)
```

Tested against the published reference over 2005-2017, that variant is
**worse**, not better. `ratio` is the median of computed over reference; the
remaining columns are the share of recipient-years within that distance:

| variant | n | ratio | median abs. diff | <0.01 | <0.05 | <0.10 |
|---|---:|---:|---:|---:|---:|---:|
| **A.** all countries, MNH = 0.15 everywhere *(this package)* | 1872 | 0.964 | 0.0245 | 35.4% | 68.2% | 83.2% |
| **B.** all countries, MNH = 0 where no burden | 1872 | 0.961 | 0.0285 | 34.8% | 56.1% | 59.1% |
| **C.** burden countries only, MNH = 0.15 | 1159 | 1.000 | 0.0072 | 56.3% | 90.7% | 95.5% |

Agreement within 0.10 falls from 83.2% to 59.1%: zeroing the constant moves a
quarter of all malaria comparisons out of even loose agreement. The reason is
visible in the zero-burden rows alone:

| | |
|---|---:|
| reference median weight, zero-burden rows | 0.2267 |
| variant A, median absolute difference | 0.0767 |
| variant B, median absolute difference | 0.2267 |
| reference weights falling **below** 0.15 | **0.0%** |

Not one of those 713 reference values is below 0.15. The published method never
assigns a malaria weight under the constant, even where its own data shows
little or no burden — so 0.15 is already *below* the reference there, and
setting it to zero opens a larger gap in the opposite direction rather than
closing one.

Variant A is therefore the package default, on the grounds that it is what
Muskoka2 does and departing would make these figures non-comparable with
published ones. There is also a substantive reading: malaria-control spending
in a malaria-free country is typically elimination maintenance, surveillance or
outbreak preparedness, and protecting pregnant women is a genuine part of that.
The constant is not obviously wrong there, merely unverifiable.

Variant B is documented rather than implemented. The two lines above are all it
takes for a sensitivity analysis, and keeping it out of the package interface
avoids offering a switch that would quietly break comparability with the
published series.

## What this vignette is not

Reference documentation. That is generated from the roxygen comments and
reached with `?sector_weights` and `?agency_weights`. A vignette explains
*when and why*; the help page explains *what and how*.
