---
title: "Getting started with rmnchfunding"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with rmnchfunding}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rmnchfunding)
```

## Why this package exists

Replace this section with the problem the package solves — the situation a
reader is in *before* they install it. A vignette that opens with a feature
list assumes the reader already knows they need one.

## A worked example

```{r}
x <- c(2, 4, 6, 8)
rescale01(x)
```

Every chunk here is executed when the package is built, so a vignette that
still knits is a second, coarser test suite: it catches interface changes
that unit tests written against internals would miss.

## What this vignette is not

Reference documentation. That is generated from the roxygen comments and
reached with `?rescale01`. A vignette explains *when and why*; the help page
explains *what and how*.
